The illusory promise of the Aligned Rank Transform

### 2025-06-27 sandbox.qmd
new version

::: {#fig-effect-main-scatterplots-1}
```{r, echo=FALSE, fig.height=3.4, message=FALSE, warning=FALSE}
source("effect_sizes_scatterplots.R")
prefix <- "6_scatter-Effect_Size-g2"
df <- readlyDataPoints(prefix,  distributions=c("norm", "lnorm", "likert5B"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)"), effect = "X1X2") %>% toCohensf2()

fig <- plotlyScatter(df, xlab = "Ground truth", ylab= "Cohen's f", xvar = "fX1X2", max = 0.6)

fig
```
Scatterplots showing the relationship between effect size estimates (Cohen's $f$) of each method and estimates of the ground-truth method for the main effect of $X_1$, when $a_1$ and $a_2$ vary within the range $[-2, 2]$. 
:::

>>>>

::: {#fig-art-binary}
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=3.2, fig.width=8}
library(gridExtra)

df <- read.csv("data-binary.csv", sep=",", header=TRUE, strip.white=TRUE)
df$S <- factor(df$S)
df$A <- factor(df$A)
df$B <- factor(df$B)

df$rnk <- rank(df$Y)

m <- mean(df$Y)
Ai <- aggregate(Y ~ A, data = df, mean)
Bj <- aggregate(Y ~ B, data = df, mean) 
AiBj <- aggregate(Y ~ A+B, data = df, mean) 
cmean <- aggregate(Y ~ A+B, data = df, mean)

library(dplyr)
df <- df %>% group_by(A) %>% mutate(Ai = mean(Y)) %>% 
  group_by(B) %>% mutate(Bj = mean(Y)) %>% 
  group_by(A,B) %>% mutate(AiBj = mean(Y)) %>% mutate(Cell = mean(Y))

df$m <- mean(df$Y)

df$artA <- df$Y - df$Cell + df$Ai - df$m
df$artB <- df$Y - df$Cell + df$Bj - df$m
df$artAB <- df$Y - df$Cell + df$AiBj - df$Ai - df$Bj + df$m

df$rnkA <- rank(round(df$artA, digits = 7))
df$rnkB <- rank(round(df$artB, digits = 7))
df$rnkAB <- rank(round(df$artAB, digits = 7))

mytheme <- gridExtra::ttheme_default(
    core = list(fg_params=list(cex = 2.0)),
    colhead = list(fg_params=list(cex = 1.0)),
    rowhead = list(fg_params=list(cex = 1.0)))

myTable <- tableGrob(df[,1:4], theme = ttheme_minimal(base_size = 10) )

diffPalette = c("#E69F00", "#FF5E00")
p1 <- (df %>%   
        ggplot(aes(x = B, y = rnk, fill = B)) + ggtitle("RNK") +
        geom_boxplot(outlier.shape = NA) +
        geom_jitter(position=position_jitter(0.3,0), size = 2, color=diffPalette[1], alpha=0.6) +
        theme_bw() + theme(legend.position = "none", plot.title = element_text(hjust = 0.5))) +  
        ylab("ranking") + scale_y_discrete(limits=as.character(1:nrow(df)))

p2 <- (df %>%   
        ggplot(aes(x = B, y = rnkB, fill = B)) + ggtitle("ART_B") +
        geom_boxplot(outlier.shape = NA) + 
        geom_jitter(position=position_jitter(0.3,0), size = 2, color=diffPalette[2], alpha=0.6) +
        theme_bw() + theme(legend.position = "none", plot.title = element_text(hjust = 0.5))) +  
        theme(axis.title.y=element_blank(), axis.text.y=element_blank()) + scale_y_discrete(limits=as.character(1:nrow(df)))

grid.arrange(myTable, p1, p2, nrow = 1)
```
Results from an experiment following a $3 \times 2$ design, where the response variable $Y$ is binary and indicates the presence of an error. The plots show the rankings produced by the simple rank transformation (RNK) and by ART for the second factor. The solid lines represent group medians.    
:::

### 2025-06-27 text/30-assumptions.qmd
new version

::: {#fig-art-binary}
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=3.2, fig.width=8}
library(gridExtra)
#library(ggplot2)

@@

diffPalette <- c("#e2e6fa", "#99b9f9")
p1 <- (df %>%   
        ggplot(aes(x = B, y = rnk, fill = B)) + ggtitle("RNK") +
        geom_boxplot(outlier.shape = NA) +
        geom_jitter(shape=1, position=position_jitter(0.3,0), size = 2) +
        theme_bw() + theme(legend.position = "none", plot.title = element_text(hjust = 0.5))) + scale_fill_manual(values=diffPalette) +  
        ylab("ranking") + scale_y_discrete(limits=as.character(1:nrow(df)))

p2 <- (df %>%   
        ggplot(aes(x = B, y = rnkB, fill = B)) + ggtitle("ART_B") +
        geom_boxplot(outlier.shape = NA) + 
        geom_jitter(shape=1, position=position_jitter(0.3,0), size = 2) +
        theme_bw() + theme(legend.position = "none", plot.title = element_text(hjust = 0.5))) + scale_fill_manual(values=diffPalette) +  
        theme(axis.title.y=element_blank(), axis.text.y=element_blank()) + scale_y_discrete(limits=as.character(1:nrow(df)))

>>>>

::: {#fig-art-binary}
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=3.2, fig.width=8}
library(gridExtra)

@@

diffPalette = c("#E69F00", "#FF5E00")
p1 <- (df %>%   
        ggplot(aes(x = B, y = rnk, fill = B)) + ggtitle("RNK") +
        geom_boxplot(outlier.shape = NA) +
        geom_jitter(position=position_jitter(0.3,0), size = 2, color=diffPalette[1], alpha=0.6) +
        theme_bw() + theme(legend.position = "none", plot.title = element_text(hjust = 0.5))) +  
        ylab("ranking") + scale_y_discrete(limits=as.character(1:nrow(df)))

p2 <- (df %>%   
        ggplot(aes(x = B, y = rnkB, fill = B)) + ggtitle("ART_B") +
        geom_boxplot(outlier.shape = NA) + 
        geom_jitter(position=position_jitter(0.3,0), size = 2, color=diffPalette[2], alpha=0.6) +
        theme_bw() + theme(legend.position = "none", plot.title = element_text(hjust = 0.5))) +  
        theme(axis.title.y=element_blank(), axis.text.y=element_blank()) + scale_y_discrete(limits=as.character(1:nrow(df)))

### 2025-06-27 text/40-interpretingEffects.qmd
new version

GLMs are defined by a link function $g$ that connects a variable —- linearly defined by the predictors -— to a response variable $Y$, which follows an arbitrary distribution. For an experimental design with two factors, $x_1$ and $x_2$, the expected value (or mean) of $Y$ conditional on $x_1$ and $x_2$ is expressed as follows:

@@

Unless we explicitely mention interpretation issues, we focus on the first scenario. Note that in the case of discrete response variables, the function $\mathcal{T}$ is monotonic but not $strictly$ monotonic, since it involves a loss of information. Consequently, an effect on the latent variable may not be observable in the responses, or the observed magnitude of the effect may appear reduced.

>>>>

GLMs are defined by a link function $g$ that connects a variable -- linearly defined by the predictors -- to a response variable $Y$, which follows an arbitrary distribution. For an experimental design with two factors, $x_1$ and $x_2$, the expected value (or mean) of $Y$ conditional on $x_1$ and $x_2$ is expressed as follows:

@@

Unless we explicitely mention interpretation issues, we focus on the first scenario. Note that in the case of discrete response variables, the function $\mathcal{T}$ is monotonic but not *strictly* monotonic, since it involves a loss of information. Consequently, an effect on the latent variable may not be observable in the responses, or the observed magnitude of the effect may appear reduced.

### 2025-06-27 text/50-ExperimentalMethod.qmd
new version

3. We evaluate three sample sizes, $n=10$, $n=20$, and $n=30$, where $n$ represents the cell size in an experimental design. For within-subjects designs where all factors are treated as repeated measures, $n$ coincides with the number $N$ of subjects (or commonly human participants in HCI research). In contrast, for a 2 $\times$ 3 between-subjects design, if the cell size is $n = 20$, then the number of subjects is $N = 120$.

4. We test the robustness of the methods under unequal variances.

5. In addition to Type I error rates, we compare the statistical power of the methods and also evaluate the quality of their effect size estimates.

6. We assess the above measures for both main and interaction effects and examine how growing effects on one or two factors affect the Type I error rate on other factors and their interactions.

@@

Previous evaluations of rank transformation methods [@Beasley:2009; @luepsen:2018] have also tested unbalanced designs, where the cell size is not constant across all levels of a factor. When combined with unequal variances, unbalanced designs are often problematic for both parametric procedures [@blanca:2018] and rank transformation methods [@Beasley:2009; @luepsen:2018]. As we mentioned earlier, we do not evaluate unbalanced designs here. However, we provide additional experimental results on missing data in the [appendix](appendix.html).

@@

We do not directly evaluate Type II errors. Instead, we report on statistical *power* defined as $Power = 1 - \beta$, where $\beta$ is the rate of Type II errors. Significance tests do not provide any power guarantees. However, we can compare the power of different methods to evaluate their relative performance. In addition to power, we assess effect size estimates, where we use as ground truth the estimates of a parametric ANOVA conducted on the latent variable $\mathcal{Y}$. While partial $\eta^2$ is the most commonly used effect size measure, we chose to base our analysis on Cohen's $f$, as its expected value is proportional to the real magnitude of effect. However, $\eta^2$ can be directly derived from Cohen's $f$ as follows:

@@

As we explained earlier, interaction effects are deformed when we transform the latent variable $\mathcal{Y}$ to obtain the observed responses. Such transformations also affect the Type I error rates and power that we observe. Here, we are interested in evaluating these measures with respect to the effects we apply to the latent variable $\mathcal{Y}$. We make this distinction whenever it is relevant in our analysis.

>>>>

3. We focus on three sample sizes, $n=10$, $n=20$, and $n=30$, where $n$ represents the cell size in an experimental design. In within-subjects designs, where all factors are treated as repeated measures, $n$ corresponds to the number of subjects $N$ (commonly human participants in HCI research). In contrast, in a 2 $\times$ 3 between-subjects design, a cell size of $n = 20$, implies a total of $N = 120$ subjects. For ordinal data only, we also report results for a larger range of sample sizes, up to $n = 512$.

4. We test the robustness of the methods when the variances on the latent variable are unequal.

5. In addition to Type I error rates, we compare the statistical power of the methods and assess the accuracy of their effect size estimates.

6. We analyze both main and interaction effects, examining how increasing the effect size of one or two factors influences the Type I error rates of other factors and their interactions.

@@

Previous evaluations of rank transformation methods [@Beasley:2009; @luepsen:2018] have also examined unbalanced designs, where cell sizes vary across the levels of a factor. When combined with unequal variances, such designs often pose challenges for both parametric procedures [@blanca:2018] and rank transformation methods [@Beasley:2009; @luepsen:2018]. As noted earlier, we do not include unbalanced designs in our evaluation. However, we provide additional experimental results on missing data in the [appendix](appendix.html).

@@

We do not directly evaluate Type II errors. Instead, we report on statistical *power* defined as $Power = 1 - \beta$, where $\beta$ is the rate of Type II errors. Significance tests do not provide any power guarantees. However, we can compare the power of different methods to evaluate their relative performance. In addition to power, we assess effect size estimates, where we use as ground truth the estimates of a parametric ANOVA conducted on the latent variable $\mathcal{Y}$. While partial $\eta^2$ is the most commonly used effect size measure, we also evaluate Cohen's $f$, as its expected value is proportional to the real magnitude of effect. However, $\eta^2$ can be directly derived from Cohen's $f$ as follows:

@@

As explained earlier, interaction effects are distorted when the latent variable $\mathcal{Y}$ is transformed to produce the observed responses. These transformations also influence the Type I error rates and statistical power we observe. In this analysis, we focus on evaluating these measures with respect to the effects applied to the latent variable $\mathcal{Y}$. We explicitly make this distinction whenever it is relevant to our discussion.

### 2025-06-27 text/65-Results-EffectSize.qmd
new version

fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab)) %>% layout(
    hovermode = "x unified",  # Shared hover on the x-axis
    yaxis = list(spikemode = "across", spikesnap = "cursor")
  )

fig

df <- data[data$effectX1 == 1,]
```
Boxplots summarizing the $\eta^2$ estimates obtained with each method for 300 data points. The population effect of $X1$ is null, and responses follow log-normal distributions. Hover over the plots to compare the methods.
:::

@@

fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab)) %>% layout(
    hovermode = "x unified",  # Shared hover on the x-axis
    yaxis = list(spikemode = "across", spikesnap = "cursor")
  )

fig

df <- data[data$effectX1 == 1,]
```
Boxplots summarizing the $\eta^2$ estimates obtained with each method for 300 data points. The population effect of $X1$ is now non-null ($a_1 = 1$). Responses follow again log-normal distributions. Hover over the plots to compare the methods.
:::

@@

>>>>

fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab)) %>% layout(
    hovermode = "x unified",  # Shared hover on the x-axis
    yaxis = list(spikemode = "across", spikesnap = "cursor"),
    xaxis = list(fixedrange=T)
  )

fig
```
Boxplots summarizing the $\eta^2$ estimates obtained with each method for 300 data points ($n=20$). The population effect of $X1$ is null, and responses follow **log-normal distributions**. Hover over the plots to compare the methods.
:::

@@

fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab)) %>% layout(
    hovermode = "x unified",  # Shared hover on the x-axis
    yaxis = list(spikemode = "across", spikesnap = "cursor"),
    xaxis = list(fixedrange=T)
  )

fig
```
Boxplots summarizing the $\eta^2$ estimates obtained with each method for 300 data points ($n=20$). The population effect of $X1$ is now non-null ($a_1 = 1$). Responses follow again **log-normal distributions**. Hover over the plots to compare the methods.
:::

@@

**Interactions**. We also provide additional results illustrating how rank-based methods can fail to accurately estimate interaction effects. @fig-eta-squared-interaction shows the estimated values of partial $\eta^2$ for the interaction term under normal distributions, as both main effects increase. Once again, we observe that RNK can substantially distort the estimated interaction. While INT reduces this distortion, the problem reappears as the main effects grow larger.

::: {#fig-eta-squared-interaction}
```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
data <- read.csv("data/effect-sizes-normal-interactions.csv", sep=",", header=TRUE, strip.white=TRUE)
data$effectX2 <- factor(data$effectX2)

ylab = "interaction - partial eta squared"
xlab = "magnitude of main effects: a1 and a2"

df <- data

fig1 <- plot_ly(df, x = ~effectX2, y = ~X1X2, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1X2, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))

fig <- subplot(fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab), xaxis = list(fixedrange=T)) %>% layout(
    hovermode = "x unified",  # Shared hover on the x-axis
    yaxis = list(spikemode = "across", spikesnap = "cursor")
  )

fig
```
Boxplots summarizing the $\eta^2$ estimates for interactions obtained with each method for 300 data points ($n=20$). The population interaction effect is null, and responses follow **normal distributions**. Hover over the plots to compare the methods.
:::

If the data take a discrete form, effect size estimates become even more distorted across all methods. @fig-eta-squared-ordinal presents results for an ordinal scale with 5 flexible levels.

::: {#fig-eta-squared-ordinal}
```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
data <- read.csv("data/effect-sizes-ordinal-interactions.csv", sep=",", header=TRUE, strip.white=TRUE)
data$effectX2 <- factor(data$effectX2)

ylab = "interaction - partial eta squared"
xlab = "magnitude of main effects: a1 and a2"

df <- data
#df <- data[data$effectX1 == 0,]

fig1 <- plot_ly(df, x = ~effectX2, y = ~X1X2, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1X2, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))

fig <- subplot(fig1, fig.par, fig.rnk, fig.int, fig.art, shareX = TRUE, shareY = T) %>% 
  layout(annotations = list(x = 0.5, y = -0.25, yshift = 0, xref = "paper", yref = "paper", xanchor = "center", yanchor = "bottom", showarrow = F, text = xlab), margin = margins) %>%
	config(displayModeBar = TRUE, scrollZoom = TRUE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale")) %>%  layout(legend = list(orientation = 'h', yanchor="bottom", xanchor="center", y = 1.2, x = .5, colors = cbPalette), yaxis = list(title = ylab), xaxis = list(fixedrange=T)) %>% layout(
    hovermode = "x unified",  # Shared hover on the x-axis
    yaxis = list(spikemode = "across", spikesnap = "cursor")
  )

fig
```
Boxplots summarizing the $\eta^2$ estimates for interactions obtained with each method for 300 data points ($n=20$). The population interaction effect is null, and responses are **ordinal with 5 flexible levels**. Hover over the plots to compare the methods.
:::

As discussed extensively earlier, such ordinal scales can distort interaction trends when main effects are non-zero, leading to interpretation issues. Although the methods may detect strong interaction effects under these conditions, the effects may not be intrinsic or meaningful.
